---
title: "Homework III"
author: "Michael Aguirre"
format:
  html: 
    toc: true # includes a table of contents in rendered html format
execute: # for all code chunks
    warning: false # do not display warnings
    message: false # do not display messages
---

**GitHub Repository Link:** <https://github.com/michaelppaull/aguirre-michael_homework-03.git>

# Setup

## Read in Packages & Data

```{r message=FALSE, warning=FALSE}
# general use
library(tidyverse)
library(readxl)
library(here)
library(janitor)
library(ggplot2)
library(dplyr)

# visualizing pairs
library(GGally)

# model selection
library(MuMIn)

# model predictions
library(ggeffects)

# model tables
library(gtsummary)
library(flextable)
library(modelsummary)

# read in data and specify path and sheet
drought_exp <- read_xlsx(path = here("data", 
                                     "Valliere_etal_EcoApps_Data.xlsx"),
                         sheet = "First Harvest")
```

## Cleaning

```{r}
# cleaning
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # adding column with species scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # moving species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # moving water_treatment column after water
```

# Problems

### Problem I

#### Define Models

```{r}
# null model
model0 <- lm(total_g ~ 1,
             data = drought_exp_clean)

# total biomass as a function of SLA, water treatment, and species
model1 <- lm(total_g ~ sla + water_treatment + species_name, 
             data = drought_exp_clean)

# total biomass as a function of SLA and water treatment
model2 <- lm(total_g ~ sla + water_treatment,
             data = drought_exp_clean)

# total biomass as a function of SLA and species
model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)

# construct model that shows total biomass as a function of water treatment and species
model4 <- lm(total_g ~ water_treatment + species,
             data = drought_exp_clean)
```

#### Compare Models

```{r}
# Create comparison table
models <- list(
  "Null" = model0,
  "Model 1" = model1,
  "Model 2" = model2,
  "Model 3" = model3,
  "Model 4" = model4
)

aic_values <- sapply(models, AIC)
bic_values <- sapply(models, BIC)
adjusted_r_squared <- sapply(models, function(model) summary(model)$adj.r.squared)

comparison <- data.frame(
  Model = names(models),
  AIC = aic_values,
  BIC = bic_values,
  Adjusted_R_Squared = adjusted_r_squared
)

# Convert to flextable
ft <- flextable(comparison)
ft <- autofit(ft)


caption <- "Table 1: Model Comparison Based on AIC, BIC, and Adjusted R-Squared Values.The table compares five models predicting total biomass (total_g) based on different predictors: SLA (Specific Leaf Area), water treatment, and species. The columns present the model names, Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC), and adjusted R-squared values for each model. Lower AIC and BIC values indicate better model performance, while higher adjusted R-squared values suggest better explanatory power."

ft <- set_caption(ft, caption)
ft
```
#### Statistical Methods
To examine the influence of SLA (Specific Leaf Area), water treatment, and species on total biomass, I fitted several linear regression models using the lm function in R. Specifically, I compared four models against the null:

1.  A model with SLA, water treatment, and species as predictors.

2.  A model with SLA and water treatment as predictors.

3.  A model with SLA and species as predictors.

4.  A model with water treatment and species as predictors.

To determine the model that best described the total biomass, I compared these models using the Akaike Information Criterion (AIC), Bayesian Information Criterion (BIC), and adjusted R-squared values. The comparison showed that Model 4, which includes water treatment and species as predictors, had the lowest AIC (-159.19595) and BIC (-138.95950) values and the highest adjusted R-squared (0.7257045) among the models tested. Therefore, Model 4 was selected as the best predictor model for total biomass.

To evaluate whether Model 4 conformed to the assumptions of a linear model, I examined diagnostic plots. The residuals vs. fitted values plot indicated that the residuals were randomly scattered around zero, suggesting linearity. The Normal Q-Q plot showed that the residuals were approximately normally distributed. The Scale-Location plot indicated homoscedasticity, as the spread of residuals was roughly constant across all levels of fitted values. Finally, the residuals vs. leverage plot did not reveal any points with excessively high leverage or influence, indicating that no single observation unduly affected the model.

Based on this analysis, Model 4, which includes water treatment and species, provides the best description of the influence of these factors on total biomass.

#### Visualize "Best" Model
```{r}
# Create a data frame with predictions
drought_exp_clean$predicted <- predict(model4, drought_exp_clean)

# generate the plot
drought_exp_clean %>% 
  ggplot(aes(x = water_treatment,
             y = total_g)) +
  geom_jitter(aes(color = water_treatment), alpha = 0.3, width = 0.2) +  # Jittered underlying data points
  geom_point(aes(y = predicted, color = water_treatment), size = 2) +
  facet_wrap(~ species_name) + # facet by species
  labs(
    title = "Model Predictions vs. Actual Data",
    x = "Water Treatment",
    y = "Total Biomass (g)"
  ) +
  scale_color_manual(values = c("Drought stressed" = "#FFD39B", # color choice for drought stressed
                                "Well watered" = "#6aa84f")) + # color choice for watered
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend
    strip.text = element_text(size = 10, face = "bold"), # adjust species name size
    axis.text.x = element_text(angle = 45, hjust = 1) # angle x-axis for easier reading
  )
```
#### Results
The predictors that best described total biomass were water treatment and species, as indicated by Model 4, which had the lowest AIC (-159), lowest BIC (-138), and highest adjusted R-squared (0.725) among the models tested. On average, well-watered plants exhibited significantly higher total biomass compared to drought-stressed plants, highlighting the substantial impact of water availability on plant growth. Additionally, significant differences were observed between species, with some species showing higher resilience and greater biomass under varying water treatments compared to others, demonstrating species-specific responses to water stress.
### Problem II
#### Affective Visualization


